---
title: "Homework7"
author: "Brendan Ryan"
date: "2025-04-07"
output: html_document
---
Brendan Ryan
bpr663
https://github.com/BrendanRyan13/SDS315HW7
```{r, echo=FALSE, message=FALSE, warning=FALSE}
setwd("C:/Users/frogb/Documents/SDS315/Homework7")
library(ggplot2)
library(tidyverse)
library(mosaic)
library(MatchIt)
```

# Problem 1: Armfolding

### A:Load and examine the data.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
ArmFoldData <- read.csv("armfold.csv")
sum(ArmFoldData$Sex == "Male")
sum(ArmFoldData$Sex == "Female")
MaleLonR <- sum(ArmFoldData$Sex == "Male" & ArmFoldData$LonR_fold == "1")/sum(ArmFoldData$Sex == "Male")
FemaleLonR <- sum(ArmFoldData$Sex == "Female" & ArmFoldData$LonR_fold == "1")/sum(ArmFoldData$Sex == "Female")
print(MaleLonR)
print(FemaleLonR)
```

There are 106 males and 111 females in the dataset. Of the males, about 47% fold their Left arm over their Right. Of the females, about 42% fold their Left arm over their Right.
### B: What is the observed difference in proportions between the two groups (males minus females)?
```{r, echo=FALSE, message=FALSE, warning=FALSE}
ObservedDiff <- MaleLonR - FemaleLonR
print(ObservedDiff)
```

The observed difference between the proportion of males and proportion of females is 0.048, or about 5%

### C: Compute a 95% confidence interval for the difference in proportions (males minus females).
```{r, echo=FALSE, message=FALSE, warning=FALSE}
se <- sqrt((MaleLonR * (1 - MaleLonR)) / 106 + (FemaleLonR * (1 - FemaleLonR)) / 111)
margin_of_error <- 1.96 * se

lower <- (MaleLonR - FemaleLonR) - margin_of_error
upper <- (MaleLonR - FemaleLonR) + margin_of_error

lower
upper

prop.test(
  x = c(sum(ArmFoldData$Sex == "Male" & ArmFoldData$LonR_fold == "1"), 
        sum(ArmFoldData$Sex == "Female" & ArmFoldData$LonR_fold == "1")),
  n = c(106, 111),
  conf.level = 0.95,
  correct = FALSE
)
```

Both the built-in R method and the calculation of the confidence interval by hand yield an interval of -0.084 - 0.180. The method to calculate the confidence interval by hand was to use this formula SE = √[ (p₁ × (1 − p₁)) / n₁ + (p₂ × (1 − p₂)) / n₂ ]. n₁ and n₂ were the number of males and females in the dataset respectively. p₁ and p₂ were the sample proportions of males who fold Left over Right and female who fold Left over Right. This resulted in me plugging in 106, 111, 0.472, and 0.423 for these variables. SE is the calculated standard error for the difference in proportions. This was used to calculated the margin of error by multiplying the standard error by a z-score of 1.96, which was selected because it would give a 95% confidence interval. 1.96 is the number of standard deviations included in the center 95% of data. 

### D: Interpret your confidence interval in context by completing the blanks in this sentence: “If we were to (blank 1),     then we would expect that (blank 2).

If we were to take a survey of the whole population, or take enough repeated samples that we sample a large majority of the population, then the difference in proportions between males and females that fold their left arm over their right arm would be between -0.084 and 0.180 with a 95% confidence.

### E: In your own words, what does the standard error you calculated above represent? What is it measuring?

The standard error is the standard deviation of the sample. It is a measure of how precise our estimated statistic is. The standard error I calculated was 0.067, which is the standard deviation of the distribution.

### F: What does the term sampling distribution refer to in this context? Be specific about, what is varying from sample      to sample, and what stays fixed.

The sampling distribution refers to the distribution of difference in proportions between males and females that would be obtained if we took many random samples from the same population. Who is in the sample varies, but the overall population and the amount sampled is fixed.

### G: What mathematical result or theorem justifies using a normal distribution to approximate the sampling distribution     of the difference in sample proportions? Explain this result briefly in your own words.

The central limit theorem is what justifies using a normal distribution in this scenario. The CLT requires a large sample size of independent events, which is true for this dataset. It is independent because one person crossing their arms a certain way has no impact on another person crossing their arms that way.

### H: Suppose your 95% confidence interval for the difference in proportions was [-0.01, 0.30]. Based on this, what would     you say to someone who claims “there’s no sex difference in arm folding”?

I would disagree. I would say that there is no way to be sure that there is no difference in arm folding, but since the 95% confidence interval includes 0, it can be reasonably said that there is no observable difference in arm folding.

### I: Imagine repeating this experiment many times with different random samples of university students. Would the           confidence interval be different across samples? Why? What should be true about the collection of allthose intervals?

The confidence interval would not be exactly the same, because of differences in sampling variability. It is up to chance if we happen to have more Left over Right in a dataset than Right over Left. In all the intervals, the 95% confidence interval is very likely to contain the true difference in proportions between men and women.

# Problem 2: Get Out The Vote

### Part A
```{r, echo=FALSE, message=FALSE, warning=FALSE}
TurnoutData <- read.csv("turnout.csv")

RecievedProportion <- sum(TurnoutData$GOTV_call==1&TurnoutData$voted1998==1)/sum(TurnoutData$GOTV_call==1)
RecievedProportion

NotRecievedProportion <- sum(TurnoutData$GOTV_call==0&TurnoutData$voted1998==1)/sum(TurnoutData$GOTV_call==0)
NotRecievedProportion

Result <- prop.test(x = c(sum(TurnoutData$GOTV_call==1&TurnoutData$voted1998==1),
                               sum(TurnoutData$GOTV_call==0&TurnoutData$voted1998==1)),
                         n = c(sum(TurnoutData$GOTV_call==1), sum(TurnoutData$GOTV_call==0)),
                         conf.level = 0.95)
Result
```

The proportion of those who received a GOTV call who voted in 1998 is 0.658. The proportion of those who did not receive a GOTV call who voted in 1998 is 0.444. We can be 95% confident that the true difference in proportions between those who voted in 1998 and received a call and those who voted in 1998 and did not receive a call lies between 0.141 and 0.266.

### Part B

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ProportionVoted1996 = prop.test(
  x = c(sum(TurnoutData$voted1996 == 1 & TurnoutData$GOTV_call == 1),
        sum(TurnoutData$voted1996 == 0 & TurnoutData$GOTV_call== 1)),
  n = c(sum(TurnoutData$voted1996 == 1), sum(TurnoutData$voted1996 == 0))
)
ProportionVoted1996

ProportionAge = prop.test(
  x = c(sum(TurnoutData$AGE > 30 & TurnoutData$GOTV_call == 1), 
        sum(TurnoutData$AGE <= 30 & TurnoutData$GOTV_call == 1)),
  n = c(sum(TurnoutData$AGE > 30), sum(TurnoutData$AGE <= 30))
)
ProportionAge

ProportionMajorParty = prop.test(
  x = c(sum(TurnoutData$MAJORPTY == 1 & TurnoutData$GOTV_call == 1), 
        sum(TurnoutData$MAJORPTY == 0 & TurnoutData$GOTV_call == 1)),
  n = c(sum(TurnoutData$MAJORPTY == 1), sum(TurnoutData$MAJORPTY == 0))
)
ProportionMajorParty

ggplot(TurnoutData) +
  geom_bar(aes(x = voted1996), fill = "maroon") +
  facet_wrap(~GOTV_call, nrow = 2, scales = "free_y") +
  labs(x = "Did the person vote in 1996? (1=yes, 0=no)",
       title = "Voted 1996 vs Voted in 1998 (1) And Didn't Vote in 1998 (0)")

ggplot(TurnoutData) +
  geom_histogram(aes(x = AGE), binwidth = 2, fill = "green") +
  facet_wrap(~GOTV_call, nrow = 2, scales = "free_y") +
  labs(x = "Age",
       title = "Ages That Voted in 1998 (1) And That Didn't Vote in 1998 (0)")

ggplot(TurnoutData) +
  geom_bar(aes(x = MAJORPTY), fill = "lightblue") +
  facet_wrap(~GOTV_call, nrow = 2, scales = "free_y") +
  labs(x = "Registered with a major U.S. political party? (1=yes, 0=no)",
       title = "MAJORPTY vs Voted in 1998 (1) And Didn't Vote in 1998 (0)")
```

Based on the graphs, I would say that all three of these variables are confounders, since there is a noticeable effect on whether or not someone voted based on if they voted in 1996, their age, and if they are registered with a major party. All of these impacts change the chance of someone voting. Older, registered, and voting experience all increase the chance of someone votes.

### Part C

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Matched <- matchit(GOTV_call ~ voted1996 + AGE + MAJORPTY,
                   data = TurnoutData,
                   ratio = 5)

summary(Matched)

MatchedData <- match.data(Matched)

TreatedProportion <- mean(MatchedData$voted1998[MatchedData$GOTV_call == 1])

ControlProportion <- mean(MatchedData$voted1998[MatchedData$GOTV_call == 0])

Result <- prop.test(
  x = c(sum(MatchedData$voted1998[MatchedData$GOTV_call == 1]), 
        sum(MatchedData$voted1998[MatchedData$GOTV_call == 0])), 
  n = c(sum(MatchedData$GOTV_call == 1), sum(MatchedData$GOTV_call == 0))
)

Result

ggplot(MatchedData) +
  geom_bar(aes(x=voted1996), fill = "maroon") +
  facet_wrap(~GOTV_call, nrow=2) +
  labs(x = "Did the person vote in the 1996 Congressional election (1=yes, 0=no)", title = "Voted 1996 vs That Voted in 1998 (1) And That Didn't Vote in 1998 (0)")
cor(TurnoutData$voted1996, TurnoutData$voted1998)

ggplot(MatchedData) +
  geom_histogram(aes(x=AGE, binwidth=2), fill = "green") +
  facet_wrap(~GOTV_call, nrow=2) +
  labs(x = "Age", title = "Ages That Voted in 1998 (1) And That Didn't Vote in 1998 (0)")
cor(TurnoutData$AGE, TurnoutData$voted1998)

ggplot(MatchedData) +
  geom_bar(aes(x=MAJORPTY), fill = "lightblue") +
  facet_wrap(~GOTV_call, nrow=2) +
  labs(x = "Is the person registerd for one of the two major U.S. political
parties (1=yes, 0=no)", title = "MAJORPTY vs That Voted in 1998 (1) And That Didn't Vote in 1998 (0)")
cor(TurnoutData$MAJORPTY, TurnoutData$voted1998)
```

The proportion of those receiving a GOTV call that voted in 1998 (after matching) is 0.648. The sample proportion of those not receiving a GOTV call who voted in 1998 (after matching) is 0.569. The 95% confidence interval between those who voted and did receive a call and those who voted and did not receive a GOTV call is 0.01 and 0.147. This means we can be 95% confidence that the true difference in proportions is between 0.01 and 0.147.

It can be concluded that there was a significant impact on voter turnout by the GOTV calls, because the confidence interval does not include 0 and the p-value is less than 0.05. Additionally, as shown by the graphs and the correlations, the matching process allows us to eliminate the confounders of age, voting experience, and party affiliation, meaning we can be more confident that a signifcant impact exists.